Income Disparity and Its Influence on Higher Education Affordability and Access\thanks{Code and data are available at: https://github.com/fanger2791/College-Tuition-and-Income-Inequality/tree/main}

### 2024-02-20 files/paper/college_tuiton_income_inequality.tex
update

---
title: "Income Disparity and Its Influence on Higher Education Affordability and Access"
author: 
  - Michael Fang
  - Harrison Huang
thanks: "Code and data are available at: https://github.com/fanger2791/College-Tuition-and-Income-Inequality/tree/main. Replication on Social Science Reproduction: https://www.socialsciencereproduction.org/reproductions/d456357c-d73d-4bce-a6ba-178bac339c7e/index"
date: today
date-format: long
abstract:   Throughout the years, we have seen how rising income inequality fuels college tuition hikes, casting a spotlight on the urgent need for policies that bridge the educational divide. This paper is a reproduction on the connection between rising income inequality and the surge in college tuition costs in the United States. Utilizing a detailed economic model, the increasing disparity in income has been a significant driver behind the growth in tuition fees since 1990, accounting for more than half of the observed increase and has also led to a decline in college attendance rates. Extending this analysis through secondary research to include Canadian universities reveals similar trends, indicating a broader North American pattern where income inequality impacts educational accessibility. This comparative approach underscores the global relevance of their findings, suggesting that addressing income disparities could play a crucial role in making higher education more accessible.
format: pdf
number-sections: true
bibliography: references.bib
---

# Introduction

In recent years, the escalating cost of college tuition in the United States has emerged as a critical issue, intertwining with the broader narrative of rising income inequality. The paper by Zhifeng Cai and Jonathan Heathcote addresses this pressing concern by examining the relationship between these two phenomena. Utilizing a competitive model of the college market, the study not only uncovers the mechanisms that link rising income inequality to increasing tuition costs but also sheds light on the consequent impacts on college attendance and social mobility [@caiandheathcote]. Through this paper, we aim to illuminate the unique challenges and opportunities facing the Canadian higher education system in the context of rising tuition costs and income inequality. Incorporating a Canadian perspective into our reproduction study adds a crucial comparative dimension, allowing us to explore the dynamics of university tuition and income inequality within a different national context. This extension acknowledges the distinct structure of the Canadian higher education system and its funding mechanisms, which differ from those in the United States in terms of government support, the magnitude of tuition fees, and the socio-economic landscape.

Central to the Zhifeng Cai and Jonathan Heathcote paper's findings is the assertion that the surge in U.S. income inequality since the 1990s can account for more than half of the observed increase in average net tuition over the same period [@caiandheathcote]. This relationship underscores a critical feedback loop where income disparities not only affect individual capacity to afford higher education but also drive institutional behaviors around tuition setting. Moreover, the paper highlights the detrimental effects of rising tuition on college attendance rates, particularly among lower-income segments, positing significant implications for social mobility and equity [@caiandheathcote].

@@

1. Recreating Table 3: We will split the table into private college versus public college. In the original work, it was not clearly stated where the data came from or for private two year studies or public four year studies. With that, we will look at data sets from the paper and recreate Table 3 from the original work. The new tables will focus on the price change from the earliest data and most recent data. In addition, the two tables are separated by private four year non profit college versus public four year college. This step includes extensive data cleaning and edits due to excel file formats. All the steps can be found in the source code to increase the reproducibility of our analysis.

2.  Recreating the Model: We will start by reproducing the competitive college market model used in the original study, ensuring our work is entirely reproducible. This step includes verifying the model's assumptions, computational methods, and data inputs.

3.  Expanding the Study with a Canadian Lens: Specifically examine how income inequality within Canada affects university tuition, comparing these dynamics with the findings from the U.S. context. This will provide insights into how different policy environments and socio-economic conditions influence the relationship between income inequality and higher education costs.

@@

# Data {#sec-data}

## Source

This paper reproduces several areas of the original paper  by Zhifeng Cai and Jonathan Heathcote [@caiandheathcote] which is originally from the American Economic Association. This reproduction utilizes the same data sets in the process. Please note that the datasets used are unfortunately not publicly accessible.

## Methodology

@@

R [@citeR] was the language and environment used for this reproduction, with different packages such as tidyverse [@tidyverse], ggplot2 [@ggplot2], readxl [@readxl], dplyr [@dplyr], tidyr [@tidyr], knitr [@knitr] packages.

## Variables

@@

1. Net Prices Data:

public_college_netprice: Data on the net prices of public colleges.

private_college_netprice: Data on the net prices of private colleges.

2. Grants Data:

public_college_grants: Data on total grants by income and year for public four-year colleges.

private_college_grants: Similar data for private four-year colleges.

3. Funding and Revenue Data:

state_funding: average state and local funding for higher education per 1000$ of personal income

public_rev_subs and private_rev_subs: detailing subsidies and revenues for public and private four-year institutions, with columns for institution type, year, net tuition revenue, subsidy, and education-related expenditures.

4. College Tuition and Fees Data:

@@

5. Income Distribution Data

@@

Density (EMG_ML_PDF): EMG_ML_PDF" stands for the Estimated Maximum Likelihood Probability Density Function. This term combines several concepts such as Estimated Maximum Likelihood (EML) which is a statistical method used to estimate the parameters of a probability distribution that make the observed data most probable. It's a common approach in econometrics and statistics for fitting models to data and Probability Density Function (PDF) which describes the likelihood of a random variable to take on a given value. In practical terms, it shows how the values of the random variable are distributed across different outcomes. This density variable shows how the probability distribution of incomes (after log transformation) changes across different income levels for the years analyzed.

# Results

```{r}
#| echo: false
#| warning: false
#| message: false

library(ggplot2)
library(readxl)
library(dplyr)
library(tidyr)
library(tibble)
library(knitr)

# Set the file path
file_path <- "/Users/michaelfang/Downloads/Figure_1_data.xlsx"

## Fig 10 is public college net prices
## Fig 11 is private college net prices 
public_college_netprice <- read_excel(file_path, sheet = "Fig 10", range = cell_cols("B:V"))
private_college_netprice <- read_excel(file_path, sheet = "Fig 11", range = cell_cols("B:V"))

## Fig 2013_12 public four year total grants by income and year
## Fig 2013_13 private four year total grants by income and year

public_college_grants <- read_excel(file_path, sheet = "Fig 2013_12", range = cell_cols("B:T"))
private_college_grants <- read_excel(file_path, sheet = "Fig 2013_13", range = cell_cols("B:T"))

# modifying column names for better access
colnames(public_college_grants) <- c("1999_low", "2003_low", "2007_low", "2011_low", "empty1", "1999_sec", "2003_sec", "2007_sec", "2011_sec", "empty2", "1999_third", "2003_third", "2007_third", "2011_third", "empty3", "1999_high", "2003_high", "2007_high", "2011_high", "empty4")

# modifying column names for better access
colnames(private_college_grants) <- c("1999_low", "2003_low", "2007_low", "2011_low", "empty1", "1999_sec", "2003_sec", "2007_sec", "2011_sec", "empty2", "1999_third", "2003_third", "2007_third", "2011_third", "empty3", "1999_high", "2003_high", "2007_high", "2011_high", "empty4")

# creating the column names
expenditure_type <- c("Net TFRB", "Grants by Income(1)", "Grants by Income(2)", "Grants by Income(3)", "Grants by Income(4)", "Funding for Higher Education per 1000$ of Income", "Tuition Revanue", "Subsidy")

## Fig 14B about the total funding in billions
#funding_for_higher_education <- read_excel(file_path, sheet = "Fig 15B")

## Fig 15A average state and local funding for higher education per 1000$ of personal income
state_funding <- read_excel(file_path, sheet = "Fig 15A")
colnames(state_funding) <- c('Year','Funding Per 1000 Income')

## Fig 17A public 4 year of subsidies and revenues in different years
public_rev_subs <- read_excel(file_path, sheet = "Fig 17A")
colnames(public_rev_subs) <- c("Institution Type","Year", "Net Tuition Revenue", "Subsidy", "Education Related Expenditures")

## Fig 17B private 4 year of subsidies and revenues in different years
private_rev_subs <- read_excel(file_path, sheet = "Fig 17B")
colnames(private_rev_subs) <- c("Institution Type","Year", "Net Tuition Revenue", "Subsidy", "Education Related Expenditures")

```

## Public Four Year College Data

```{r}
#| label: tbl-public_data
#| tbl-cap: Public Four Year College Data
#| warning: false
#| message: false
#| echo: false

# this is to shorten the name allowing cleaner code and set all items to numeric
pcn <- public_college_netprice
pcg <- public_college_grants

# rounding numbers to get rid of long decimals
a <- round(as.numeric(pcg$"1999_low"[13]), 0)
a1 <- round(as.numeric(pcg$"1999_sec"[13]), 0)
a2 <-round(as.numeric(pcg$"1999_third"[13]), 0)
a3 <-round(as.numeric(pcg$"1999_high"[13]), 0)
a4 <-round(as.numeric(state_funding$`Funding Per 1000 Income`[1]), 3)

# creating column of data1
pub_earliest_data <-c(pcn$"96-97"[4], a, a1, a2, a3, a4, public_rev_subs$"Net Tuition Revenue"[2], public_rev_subs$"Subsidy"[2])

b <- round(as.numeric(pcg$"2011_low"[13]), 0)
b1 <- round(as.numeric(pcg$"2011_sec"[13]), 0)
b2 <- round(as.numeric(pcg$"2011_third"[13]), 0)
b3 <- round(as.numeric(pcg$"2011_high"[13]), 0)
b4 <- round(as.numeric(state_funding$`Funding Per 1000 Income`[31]), 3)

pub_latest_data <-c(public_college_netprice$"16-17"[4],  b, b1, b2, b3, b4, public_rev_subs$"Net Tuition Revenue"[4], public_rev_subs$"Subsidy"[4])

pub_percent_change <-c("62.77%", "68.21%", "80.53%", "69.74%", "78.52%", "-28.31%", "47.35%", "-8.5%")

pub_year_frame <-c("1996 to 2016", "1999 to 2011","1999 to 2011", "1999 to 2011","1999 to 2011","1984 to 2014", "2003 to 2013", "2003 to 2013")

# create frame for public data
public_data <- tibble("Type" = expenditure_type, "Data_1" = pub_earliest_data, "Data_2" = pub_latest_data, "PercentChange" = pub_percent_change, "TimeFrame" = pub_year_frame)

# create the table
public_data_table <-
head (public_data, 8) |>
  kable(
  booktabs = TRUE
  )

public_data_table
```
## Private Four Year College Data

```{r}
#| label: tbl-private_data
#| tbl-cap: Private Four Year College Data
#| warning: false
#| message: false
#| echo: false

prcn <- private_college_netprice
prcg <- private_college_grants

# rounding data and making sure that number is not too long
c <- round(as.numeric(prcg$"1999_low"[4]), 0)
c1 <- round(as.numeric(prcg$"1999_sec"[4]), 0)
c2 <-round(as.numeric(prcg$"1999_third"[4]), 0)
c3 <-round(as.numeric(prcg$"1999_high"[4]), 0)
c4 <-round(as.numeric(state_funding$`Funding Per 1000 Income`[1]), 3)

# creating data 1 column
pri_earliest_data <-c(prcn$"96-97"[4],c, c1, c2, c3, c4, private_rev_subs$"Net Tuition Revenue"[2], private_rev_subs$"Subsidy"[2])

# rounding data and making sure that number is not too long
d <- round(as.numeric(prcg$"2011_low"[4]), 0)
d1 <- round(as.numeric(prcg$"2011_sec"[4]), 0)
d2 <-round(as.numeric(prcg$"2011_third"[4]), 0)
d3 <-round(as.numeric(prcg$"2011_high"[4]), 0)
d4 <-round(as.numeric(state_funding$`Funding Per 1000 Income`[31]), 3)

pri_latest_data <-c(prcn$"16-17"[4], d, d1,d2, d3, d4, private_rev_subs$"Net Tuition Revenue"[4], private_rev_subs$"Subsidy"[4])

pri_percent_change <-c("30.27%", "120%", "80.84%", "28.30%", "97.36%", "-28.31%", "17.62%", "28.86%")

pri_year_frame <-c("1996 to 2016", "1999 to 2011","1999 to 2011", "1999 to 2011","1999 to 2011","1984 to 2014", "2003 to 2013", "2003 to 2013")

# create frame for private data
private_data <- data.frame("Type" = expenditure_type, "Data_1" = pri_earliest_data, "Data_2" = pri_latest_data, "PercentChange" = pri_percent_change, "YearFrame" = pri_year_frame)

# create the table
private_data_table <-
head (private_data, 8) |>
  kable(
  booktabs = TRUE
  )
private_data_table

```

In the two tables @tbl-public_data and @tbl-private_data, the first column stands for tuition, fees, room and board. This is a summary data of the entire expense at that specific institution. The Grants by Income data then looks grants by year given to each income group. For example, the second row of the table looks at grants given out to people that has an annual income of less than 30,000. The sixth row looks at state funding according to an individual's income by per 1000. For example, if an individual were to make 50,000 a year, state funding for higher education to that individual should be 368.5$. There was no private and public data, as a result the state funding data is the same for both private and public colleges. The last two row looks at the data of the institution's revenue and expenditure on subsidizing students. The year frame was different for each data sets, therefore, we decided to create a separate column for cleaner look for the table. This would prevent N/A cells and could be confusing for readers.

First, looking at @tbl-public_data, there is a big increase in the net fees at 62.77% increase in just 20 years. Then looking at the grants given out by the public institutions, even though we see a continuous increase of more funding for the two lower income groups, we are still seeing a big increase in percentage for higher income individuals. When looking at state funding, there is even a -28.31 % decrease in the funding for higher education. Lastly looking at the tuition revenue and subsidy, despite making steady profits, the institution on the other hand decreased subsitizing expenditures by 8.5% from 2003 to 2014.

Second, here are some key findings from @tbl-private_data. There is a less increase of fees at 30.27%. However, the total expenditure is significant more compared to public institutions. Then looking at the grants given out by private institutions, with the lowest income group gaining a significant increase of 120%. Interestingly, the highest group also gained 97.36%. In the same situation, state funding, showed a -28.31 % decrease for higher education. Lastly looking at the tuition revenue and subsidy, 17.62% increase in tuition revenue, and also an increase of 28.86% subsidy expenditures.

@@

## College Tuition And Fees

```{r}
#| label: fig-stickers
#| fig-cap: College Tuition And Fees, US$(2016)
#| echo: false
#| warning: false

# Load necessary libraries
library(ggplot2)
library(readxl)
library(dplyr)
library(tidyr)
library(readr)

# Set the file path 
file_path <- "../data/analysis_data/college_tuition_fees_analysis_data.csv"

# Read the data from the CSV file
data <- read_csv(file_path)

# Convert data types if necessary - all price columns should be numeric
data <- mutate(data, 
               `Public Sticker` = as.numeric(`Public Sticker`),
               `Public Net` = as.numeric(`Public Net`),
               `Private Sticker` = as.numeric(`Private Sticker`),
               `Private Net` = as.numeric(`Private Net`))

# Remove rows with NA values in any of the price columns
data <- na.omit(data)

# Ensure 'Academic Year' is a factor to maintain order in the plot
data$`Academic Year` <- factor(data$`Academic Year`, levels = unique(data$`Academic Year`))

# Reshape data to long format for ggplot2
data_long <- gather(data, key = "Price Type", value = "Amount", -`Academic Year`)

# Plot using the reshaped data
p <- ggplot(data_long, aes(x=`Academic Year`, y=Amount, group=`Price Type`, colour=`Price Type`)) +
  geom_line() +
  labs(
       x='Academic Year', y='Price in $') +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  scale_colour_manual(values = c("Public Sticker"="blue", "Public Net"="green", "Private Sticker"="red", "Private Net"="purple"))

# Print the plot to see if it appears in the R console
print(p)

# Save the plot
ggsave("figure_1_plot.png", plot = p, width = 12, height = 8)
```

@fig-stickers, as presented, showcases a multi-decade overview of college tuition and fees in the United States, providing a clear visual representation of trends across both public and private institutions, accounting for inflation to 2016 dollars. The graph is particularly telling in its illustration of the divergent paths between sticker prices and net prices over time.

@@

## Comparison of Income Distribution for 1989 and 2016

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  letterpaper,
  DIV=11,
  numbers=noendperiod]{scrartcl}

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{5}
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
% definitions for citeproc citations
\NewDocumentCommand\citeproctext{}{}
\NewDocumentCommand\citeproc{mm}{%
  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
\makeatletter
 % allow citations to break across lines
 \let\@cite@ofmt\@firstofone
 % avoid brackets around text for \cite:
 \def\@biblabel#1{}
 \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
 {\begin{list}{}{%
  \setlength{\itemindent}{0pt}
  \setlength{\leftmargin}{0pt}
  \setlength{\parsep}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
   \setlength{\leftmargin}{\cslhangindent}
   \setlength{\itemindent}{-1\cslhangindent}
  \fi
  % set entry spacing
  \setlength{\itemsep}{#2\baselineskip}}}
 {\end{list}}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\KOMAoption{captions}{tableheading}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Income Disparity and Its Influence on Higher Education Affordability and Access},
  pdfauthor={Michael Fang; Harrison Huang},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{Income Disparity and Its Influence on Higher Education
Affordability and Access\thanks{Code and data are available at:
https://github.com/fanger2791/College-Tuition-and-Income-Inequality/tree/main.
Replication on Social Science Reproduction:
https://www.socialsciencereproduction.org/reproductions/d456357c-d73d-4bce-a6ba-178bac339c7e/index}}
\author{Michael Fang \and Harrison Huang}
\date{February 19, 2024}

\begin{document}
\maketitle
\begin{abstract}
Throughout the years, we have seen how rising income inequality fuels
college tuition hikes, casting a spotlight on the urgent need for
policies that bridge the educational divide. This paper is a
reproduction on the connection between rising income inequality and the
surge in college tuition costs in the United States. Utilizing a
detailed economic model, the increasing disparity in income has been a
significant driver behind the growth in tuition fees since 1990,
accounting for more than half of the observed increase and has also led
to a decline in college attendance rates. Extending this analysis
through secondary research to include Canadian universities reveals
similar trends, indicating a broader North American pattern where income
inequality impacts educational accessibility. This comparative approach
underscores the global relevance of their findings, suggesting that
addressing income disparities could play a crucial role in making higher
education more accessible.
\end{abstract}

\section{Introduction}\label{introduction}

In recent years, the escalating cost of college tuition in the United
States has emerged as a critical issue, intertwining with the broader
narrative of rising income inequality. The paper by Zhifeng Cai and
Jonathan Heathcote addresses this pressing concern by examining the
relationship between these two phenomena. Utilizing a competitive model
of the college market, the study not only uncovers the mechanisms that
link rising income inequality to increasing tuition costs but also sheds
light on the consequent impacts on college attendance and social
mobility (Cai and Heathcote 2022). Through this paper, we aim to
illuminate the unique challenges and opportunities facing the Canadian
higher education system in the context of rising tuition costs and
income inequality. Incorporating a Canadian perspective into our
reproduction study adds a crucial comparative dimension, allowing us to
explore the dynamics of university tuition and income inequality within
a different national context. This extension acknowledges the distinct
structure of the Canadian higher education system and its funding
mechanisms, which differ from those in the United States in terms of
government support, the magnitude of tuition fees, and the
socio-economic landscape.

Central to the Zhifeng Cai and Jonathan Heathcote paper's findings is
the assertion that the surge in U.S. income inequality since the 1990s
can account for more than half of the observed increase in average net
tuition over the same period (Cai and Heathcote 2022). This relationship
underscores a critical feedback loop where income disparities not only
affect individual capacity to afford higher education but also drive
institutional behaviors around tuition setting. Moreover, the paper
highlights the detrimental effects of rising tuition on college
attendance rates, particularly among lower-income segments, positing
significant implications for social mobility and equity (Cai and
Heathcote 2022).

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Recreating Table 3: We will split the table into private college
  versus public college. In the original work, it was not clearly stated
  where the data came from or for private two year studies or public
  four year studies. With that, we will look at data sets from the paper
  and recreate Table 3 from the original work. The new tables will focus
  on the price change from the earliest data and most recent data. In
  addition, the two tables are separated by private four year non profit
  college versus public four year college. This step includes extensive
  data cleaning and edits due to excel file formats. All the steps can
  be found in the source code to increase the reproducibility of our
  analysis.
\item
  Recreating the Model: We will start by reproducing the competitive
  college market model used in the original study, ensuring our work is
  entirely reproducible. This step includes verifying the model's
  assumptions, computational methods, and data inputs.
\item
  Expanding the Study with a Canadian Lens: Specifically examine how
  income inequality within Canada affects university tuition, comparing
  these dynamics with the findings from the U.S. context. This will
  provide insights into how different policy environments and
  socio-economic conditions influence the relationship between income
  inequality and higher education costs.
\end{enumerate}

@@

\section{Data}\label{sec-data}

\subsection{Source}\label{source}

This paper reproduces several areas of the original paper by Zhifeng Cai
and Jonathan Heathcote (Cai and Heathcote 2022) which is originally from
the American Economic Association. This reproduction utilizes the same
data sets in the process. Please note that the datasets used are
unfortunately not publicly accessible.

\subsection{Methodology}\label{methodology}

@@

R (R Core Team 2023) was the language and environment used for this
reproduction, with different packages such as tidyverse (Wickham et al.
2019), ggplot2 (Wickham 2016), readxl (Wickham and Bryan 2023), dplyr
(Wickham et al. 2023), tidyr (Wickham et al. 2024), knitr (Xie 2023)
packages.

\subsection{Variables}\label{variables}

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Net Prices Data:
\end{enumerate}

public\_college\_netprice: Data on the net prices of public colleges.

private\_college\_netprice: Data on the net prices of private colleges.

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{1}
\tightlist
\item
  Grants Data:
\end{enumerate}

public\_college\_grants: Data on total grants by income and year for
public four-year colleges.

private\_college\_grants: Similar data for private four-year colleges.

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{2}
\tightlist
\item
  Funding and Revenue Data:
\end{enumerate}

state\_funding: average state and local funding for higher education per
1000\$ of personal income

public\_rev\_subs and private\_rev\_subs: detailing subsidies and
revenues for public and private four-year institutions, with columns for
institution type, year, net tuition revenue, subsidy, and
education-related expenditures.

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{3}
\tightlist
\item
  College Tuition and Fees Data:
\end{enumerate}

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{4}
\tightlist
\item
  Income Distribution Data
\end{enumerate}

@@

Density (EMG\_ML\_PDF): EMG\_ML\_PDF'' stands for the Estimated Maximum
Likelihood Probability Density Function. This term combines several
concepts such as Estimated Maximum Likelihood (EML) which is a
statistical method used to estimate the parameters of a probability
distribution that make the observed data most probable. It's a common
approach in econometrics and statistics for fitting models to data and
Probability Density Function (PDF) which describes the likelihood of a
random variable to take on a given value. In practical terms, it shows
how the values of the random variable are distributed across different
outcomes. This density variable shows how the probability distribution
of incomes (after log transformation) changes across different income
levels for the years analyzed.

\section{Results}\label{results}

\subsection{Public Four Year College
Data}\label{public-four-year-college-data}

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.5444}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.0778}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.0778}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1556}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1444}}@{}}

\caption{\label{tbl-public\_data}Public Four Year College Data}

\tabularnewline

\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Type
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Data\_1
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Data\_2
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
PercentChange
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
TimeFrame
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Net TFRB & 8730 & 14210 & 62.77\% & 1996 to 2016 \\
Grants by Income(1) & 5847 & 9835 & 68.21\% & 1999 to 2011 \\
Grants by Income(2) & 3693 & 6667 & 80.53\% & 1999 to 2011 \\
Grants by Income(3) & 1748 & 2967 & 69.74\% & 1999 to 2011 \\
Grants by Income(4) & 1443 & 2576 & 78.52\% & 1999 to 2011 \\
Funding for Higher Education per 1000\$ of Income & 7.37 & 5.283 &
-28.31\% & 1984 to 2014 \\
Tuition Revanue & 6610 & 9740 & 47.35\% & 2003 to 2013 \\
Subsidy & 8350 & 7640 & -8.5\% & 2003 to 2013 \\

\end{longtable}

\subsection{Private Four Year College
Data}\label{private-four-year-college-data}

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.5444}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.0778}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.0778}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1556}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1444}}@{}}

\caption{\label{tbl-private\_data}Private Four Year College Data}

\tabularnewline

\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Type
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Data\_1
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Data\_2
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
PercentChange
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
YearFrame
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Net TFRB & 20020 & 26080 & 30.27\% & 1996 to 2016 \\
Grants by Income(1) & 12155 & 22827 & 120\% & 1999 to 2011 \\
Grants by Income(2) & 11821 & 20355 & 80.84\% & 1999 to 2011 \\
Grants by Income(3) & 10333 & 15389 & 28.30\% & 1999 to 2011 \\
Grants by Income(4) & 6328 & 12489 & 97.36\% & 1999 to 2011 \\
Funding for Higher Education per 1000\$ of Income & 7.37 & 5.283 &
-28.31\% & 1984 to 2014 \\
Tuition Revanue & 19860 & 23360 & 17.62\% & 2003 to 2013 \\
Subsidy & 15560 & 20050 & 28.86\% & 2003 to 2013 \\

\end{longtable}

In the two tables Table~\ref{tbl-public_data} and
Table~\ref{tbl-private_data}, the first column stands for tuition, fees,
room and board. This is a summary data of the entire expense at that
specific institution. The Grants by Income data then looks grants by
year given to each income group. For example, the second row of the
table looks at grants given out to people that has an annual income of
less than 30,000. The sixth row looks at state funding according to an
individual's income by per 1000. For example, if an individual were to
make 50,000 a year, state funding for higher education to that
individual should be 368.5\$. There was no private and public data, as a
result the state funding data is the same for both private and public
colleges. The last two row looks at the data of the institution's
revenue and expenditure on subsidizing students. The year frame was
different for each data sets, therefore, we decided to create a separate
column for cleaner look for the table. This would prevent N/A cells and
could be confusing for readers.

First, looking at Table~\ref{tbl-public_data}, there is a big increase
in the net fees at 62.77\% increase in just 20 years. Then looking at
the grants given out by the public institutions, even though we see a
continuous increase of more funding for the two lower income groups, we
are still seeing a big increase in percentage for higher income
individuals. When looking at state funding, there is even a -28.31 \%
decrease in the funding for higher education. Lastly looking at the
tuition revenue and subsidy, despite making steady profits, the
institution on the other hand decreased subsitizing expenditures by
8.5\% from 2003 to 2014.

Second, here are some key findings from Table~\ref{tbl-private_data}.
There is a less increase of fees at 30.27\%. However, the total
expenditure is significant more compared to public institutions. Then
looking at the grants given out by private institutions, with the lowest
income group gaining a significant increase of 120\%. Interestingly, the
highest group also gained 97.36\%. In the same situation, state funding,
showed a -28.31 \% decrease for higher education. Lastly looking at the
tuition revenue and subsidy, 17.62\% increase in tuition revenue, and
also an increase of 28.86\% subsidy expenditures.

@@

\subsection{College Tuition And Fees}\label{college-tuition-and-fees}

\begin{figure}

\centering{

\includegraphics{college_tuiton_income_inequality_files/figure-pdf/fig-stickers-1.pdf}

}

\caption{\label{fig-stickers}College Tuition And Fees, US\$(2016)}

\end{figure}%

Figure~\ref{fig-stickers}, as presented, showcases a multi-decade
overview of college tuition and fees in the United States, providing a
clear visual representation of trends across both public and private
institutions, accounting for inflation to 2016 dollars. The graph is
particularly telling in its illustration of the divergent paths between
sticker prices and net prices over time.

@@

\subsection{Comparison of Income Distribution for 1989 and
2016}\label{comparison-of-income-distribution-for-1989-and-2016}
